A Formal Theory of Bidirectional Typing for Simple Types

### 2023-05-30 tex/BiSig.tex
[ update ] A seperate subsection on Annotatability

\subsubsection{Language formalization and its frameworks}
\cite{Wadler2022}
PoplMark Challenge~\citep{Aydemir2005}

@@

\section{Bidirectional simply typed \texorpdfstring{$\lambda$}{lambda}-calculus}\label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

\subsection{Type Synthesis and Checking via Bidirectional Typing}
\subsection{Formalisation in \Agda}

@@

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
  In our setting, the set of $\Sigma$-types over $\Theta$ may contain variables from $\Theta$, so $\Type_{\Sigma}(\Theta)$ is non-empty if $\Theta$ is.
  Therefore, there is no need to introduce base types that are not used by any term construct.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists solely of a binary operation $\bto$, where $\arity(\bto) = 2$.
  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
\end{example}

@@

\subsection{Binding Signatures}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $\Xi$ is the number of type variables, 
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \vdash (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

\begin{example} \label{ex:STLC-sig}
  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B), A \to B && \mathsf{abs}\colon A , B \vdash [A]B \to (A \bto B)
  \end{align*}
    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
  \begin{align*}
    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
    && 
    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
  \end{align*}
  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
to $\Omega_{\bto}$:
  \begin{align*}
    \mathsf{pair}\colon A, B \vdash A, B \to A \btimes B
    && \fst \colon A, B \vdash A \btimes B \to A
    && \snd \colon A, B \vdash A \btimes B \to B
  \end{align*}
\end{example}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \mid \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
    and $\rho : \Sub{\Xi}{\emptyset}$}
  \end{mathpar}
  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
  \label{fig:intrinsic-typing}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
    \\
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{\Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \vdash [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
    and $\rho : \Sub{\Xi}{\emptyset}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \Mode \times \Type_{\Sigma}(\Xi)\right)^* \times \Mode \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

  For brevity, we write $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
\end{definition}

\begin{example}
  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \vdash (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \vdash [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
\end{example}

\subsection{Bidirectionally Typed Terms: Intrinsic Typing vs.\ Extrinsic Typing}

\subsubsection{Intrinsically Typed Terms}

\subsubsection{Extrinsically Typed Terms}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}

  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}\;d$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \vdash [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

@@

\subsection{Between Binding Signatures and their Bidirectional Variants}
\subsubsection{Soundness}
\subsubsection{Completeness (Annotatability)}
\subsection{Mode Correctness}

@@

\subsection{Substitution Properties}
\subsection{Generic First-Order Unification}
\cite{McBride2003,McBride2003a}
\begin{definition}
A \emph{decidable equality} for a type $X$ is a dependent function $\prod_{x, y : X} \left(\id{x}{y}\right) + \neg \left(\id{x}{y}\right)$.
\end{definition}

@@

\subsection{Type Synthesis and Checking with Unification}

\section{Formal Implementation and Examples}
\begin{LTannote}{de Bruijn}
  In our Agda formalisation, we use de Bruijn indices to represent variables, so we don't use an arbitrary type $\Theta$ but the number of variables as the index of typing judgements.
\end{LTannote}

\citep{Xie2018}

>>>>

\citep{Xie2018}
\subsubsection{Language formalization and its frameworks}
\cite{Wadler2022}
PoplMark Challenge~\citep{Aydemir2005}

@@

\section{A Case Study: Bidirectional Simply Typed \texorpdfstring{$\lambda$}{lambda}-Calculus}\label{sec:bi-stlc}
\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash \isTerm{t} \Rightarrow A}
    \quad\text{and}\quad \boxed{\Gamma \vdash \isTerm{t} \Leftarrow A}
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash \isTerm{x} \Rightarrow A}\;(\text{var}^\Rightarrow)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Leftarrow B \\ A = B}{\Gamma \vdash \isTerm{t : B}\Rightarrow A}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow B \\ A = B}{\Gamma \vdash \isTerm{t\uparrow} \Leftarrow A}\;(\text{sub})
    \and
    \inferrule{\Gamma \vdash \isTerm{t} \Rightarrow A \bto B \\ \Gamma \vdash \isTerm{u} \Leftarrow A}{\Gamma \vdash \isTerm{t\;u} \Rightarrow B}
    \and
    \inferrule{\Gamma, \isTerm{x} : A \vdash \isTerm{t} \Leftarrow B}{\Gamma \vdash \isTerm{\lam{x}t} \Leftarrow A \bto B}
  \end{mathpar}
  \caption{The extrinsic style of bidirectional simply typed $\lambda$-calculus with type variables}
  \label{fig:bi-stlc}
\end{figure}
\begin{remark}
  We avoid using the term `function type' and its conventional notation $\to$ at the object level on purpose, as it may be confused with function types in our type-theoretic meta-language.
\end{remark}

@@

\begin{example}\label{ex:implication}
  Simply typed $\lambda$-calculus typically includes a binary implication type denoted by $\bto$ and some base types to ensure that the set of all types is non-empty.
  The type signature $\Sigma_{\bto}$ of simply typed $\lambda$-calculus consists of a binary operation $\bto$, where $\arity(\bto) = 2$ and an operation $N$ of arity $0$.
  In the case of simply typed $\lambda$-calculus with binary products, we can extend $\Sigma_{\bto}$ by adding a binary operation $\btimes$ to represent the binary product type.
  This extended signature is denoted as $\Sigma_{\bto, \btimes}$.
\end{example}

@@

\subsection{Binding Signatures}
\begin{definition}\label{def:binding-signature}
  For a type signature $\Sigma$, a \emph{binding signature} $\Omega$ consists of a set $O$ and a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_\Sigma(\Xi) \times \Type_\Sigma(\Xi)\right)^* \times \Type_\Sigma(\Xi).
  \]
  Each inhabitant $o: O$ is meant to represent a term construct $\tmOp_o$ in a simple type theory with a triple $\left(\Xi, \left[\left(\Delta_1; A_1\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right], A\right)$
  instead of a number as its arity $\arity_o$ where
  \begin{enumerate}
    \item $\Xi$ is the number of type variables, 
    \item $A : \Type_\Sigma(\Xi)$ is the target type of $\tmOp_o$, and
    \item $\left[\left(\Delta_1; A_{1}\right), \ldots, \left(\Delta_{n}; A_{n}\right) \right]$ is a list of pairs where
    \item the $i$-th pair $(\Delta_i, A_i) :\Cxt_{\Sigma}(\Xi) \times \Term_{\Sigma}(\Xi)$ are types of the binding variables and the type of the $i$-th argument of $\tmOp_{o}$.
  \end{enumerate}
  For brevity, we write $o \colon \Xi \rhd (\Delta_1)A_{1}, \ldots, \left[\Delta_{n}\right] A_{n} \to A$ to indicate an operation $o$ with its arity. 
\end{definition}
\begin{remark}[Terminology on binding signature]
  \cite{Aczel1978,Fiore2010}
\end{remark}

\begin{example} \label{ex:STLC-sig}
  Given the type signature $\Sigma_{\bto}$ for the implication type, the term signature $\Lambda_{\bto}$ for simply typed $\lambda$-calculus can be described by operations
  \begin{align*}
    \mathsf{app}\colon A, B \rhd (A \bto B), A \to B && \mathsf{abs}\colon A , B \rhd [A]B \to (A \bto B)
  \end{align*}
    or, verbosely, the signature $\Lambda_{\bto}$ consists of the type $O_\Lambda = \{\app, \abs\}$ with arities
  \begin{align*}
    \arity(\app) = (\{A, B\}, [(\cdot; A \bto B), (\cdot; A)], B)
    && 
    \arity(\abs) = (\{A, B\}, [(\cdot, A; B)], A \bto B)
  \end{align*}
  For the type signature $\Sigma_{\bto, \btimes}$, the term signature $\Omega_{\bto, \btimes}$ for simply typed $\lambda$-calculus with finite products has three additional operations
to $\Omega_{\bto}$:
  \begin{align*}
    \mathsf{pair}\colon A, B \rhd A, B \to A \btimes B
    && \fst \colon A, B \rhd A \btimes B \to A
    && \snd \colon A, B \rhd A \btimes B \to B
  \end{align*}
\end{example}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} A} \\
    \inferrule{A \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} A}\;(\var)
    \and
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset}  \\ \Gamma, \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$}
  \end{mathpar}
  \caption{Intrinsic typing rules for a simple type theory $(\Sigma, \Omega)$}
  \label{fig:intrinsic-typing}
\end{figure}

@@

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}}\;(\text{anno})
    \\
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(t_1, \ldots, t_n)}}\;(\text{op}) 
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Annotatable raw terms for $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A} \quad \text{where $\vdash_{\Sigma, \Omega} t$} \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A}\;(\var)
    \and
    \inferrule{\Gamma \vdash \isTerm{t} : A}{\Gamma \vdash (\isTerm{t \annote A}) : A}\;(\text{anno})
    \and
    \inferrule{\Gamma, \isTerm{\vec{x}_1} : \simsub{\Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_1} : \simsub{A_{1}}{\rho} \quad\cdots\quad \Gamma, \isTerm{\vec{x}_n} : \simsub{\Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}}
    {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\,t_1; \ldots; \vec{x}_n.\,t_n)} : \simsub{A}{\rho}}\;(\tmOp)
    \and \text{for $o\colon \Xi \rhd [\Delta_1]A_1, \ldots, [\Delta_{n}]A_{n} \to A$ in $\Omega$
    and $\rho : \Sub{\Xi}{\emptyset}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a simple type theory $(\Sigma, \Omega)$ with annotation}
  \label{fig:extrinsic-typing}
\end{figure}

@@

\begin{definition}
  For a type signature $\Sigma$, a \emph{bidirectional binding signature} $\Omega$ is a set $O$ with a function
  \[
    \mathit{ar}\colon O \to \sum_{\Xi : \N} \left(\Cxt_{\Sigma}(\Xi) \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi)\right)^* \times \isDir{\Mode} \times \Type_{\Sigma}(\Xi).
  \]
  where $\Mode$ consists of $\chk$ for checking and $\syn$ for synthesis.
  Bidirectional binding signatures are just binding signatures (\Cref{def:binding-signature}) augmented with a mode for each argument and its target of a construct $\tmOp_o$ in a bidirectional system, i.e.
  an arity is a $4$-tuple
  \[
    \left(\Xi, \left[\left(\Delta_1, d_1, A_1\right), \ldots, \left(\Delta_{n}, d_n, A_{n}\right) \right], d, A\right)
  \]
  where $d$ and $d_i$'s indicate the modes of a construct and its arguments respectively.

  For brevity, we write $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ to indicate an operation $o$ with its arity. 
\end{definition}

\begin{example}
  The term signature $\Lambda_{\bto}^{\leftrightarrows}$ for bidirectional simply typed $\lambda$-calculus introduced in \Cref{sec:bi-stlc} can be specified by operations 
  \begin{align*}
    \mathsf{app}\colon A, B \rhd (A \bto B)^{\Rightarrow}, A^{\Leftarrow} \to B^{\Rightarrow} &&
    \mathsf{abs}\colon A , B \rhd [A]B^{\Leftarrow} \to (A \bto B)^{\Leftarrow}
  \end{align*}
  extending $\Lambda_{\bto}$ (\Cref{ex:STLC-sig}) with the mode information.
\end{example}

\begin{LTannote}{deriving from annotation}
  deriving bidirectional binding signature from annotation
\end{LTannote}
\begin{theorem}[Signature annotation]

\end{theorem}

\subsection{Extrinsically Typed Terms for Bidirectional Type Systems}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{d}}
    \\
    \inferrule{x : \Identifier}{\vdash_{\Sigma, \Omega} \isTerm{x}\;\isDir{\syn}}\;(\text{var})
    \and
    \inferrule{\cdot \vdash_{\Sigma} A \\ \vdash_{\Sigma, \Omega}\isTerm{t}\;\isDir{\chk}}{\vdash_{\Sigma, \Omega} \isTerm{t \annote A}\;\isDir{\syn}}\;(\text{anno})
    \and
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t}\;\isDir{\syn}}{\vdash_{\Sigma, \Omega} (\isTerm{t \uparrow})\;\isDir{\chk}}
  \end{mathpar}
  \begin{mathpar}
    \inferrule{\vdash_{\Sigma, \Omega} \isTerm{t_1}\;\isDir{d_1} \quad \cdots \quad \vdash_{\Sigma, \Omega} \isTerm{t_n}\;\isDir{d_n}}
    {\vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1.\, t_1; \ldots;\vec{x}_n.\, t_n)}\;\isDir{d}}\;(\text{op})
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

\begin{figure}
  \centering
  \small
  \begin{mathpar}
    \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\isDir{d}}
    \quad \text{where $\vdash_{\Sigma, \Omega} \isTerm{t}\;d$} 
    \\
    \inferrule{(x : A) \in \Gamma}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{x} : A^\syn}\;(\var)
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^\chk \\ A = B}{\mid \Gamma \vdash_{\Sigma, \Omega} (\isTerm{t \annote B}):  A^\syn}\;(\text{anno})
    \and
    \inferrule{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : B^{\syn} \\ A = B}{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t\uparrow} : A^\chk}\;(\text{sub})
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\isTerm{\vec{x}_{1}} : \Delta_{1}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_{1}} : \simsub{A_{1}}{\rho}^{\isDir{d_1}} \\
      \cdots \\
    \Gamma, \simsub{\isTerm{\vec{x}_{n}} : \Delta_{n}}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_n} : \simsub{A_{n}}{\rho}^{\isDir{d_n}}}
  {\Gamma \vdash_{\Sigma, \Omega} \isTerm{\tmOp_o(\vec{x}_1 .\, t_{1}; \ldots; \vec{x}_{n}.\, t_{n})} : \simsub{A}{\rho}^{\isDir{d}}} \; (\tmOp)
    \and \text{for $o \colon \Xi \rhd [\Delta_1]A_{1}^{d_1}, \ldots, [\Delta_{n}] A^{d_n}_{n} \to A^{d}$ in $\Omega$}
    \\
    \text{Abbreviations: $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Rightarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\syn}$ and $\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \Leftarrow A} \defeq \boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} : A^\chk}$}
  \end{mathpar}
  \caption{Extrinsic typing rules for a bidirectional system $(\Sigma, \Omega)$}
\end{figure}

@@

\subsection{Soundness and Completeness with respect to Binding Signatures}

\subsection{Annotatability}

@@

\subsection{Mode Correctness}
%\subsection{Substitution Properties}
%\subsection{Generic First-Order Unification}
%\cite{McBride2003,McBride2003a}

@@

\subsection{Type Synthesis and Checking}

\section{Formalisation in \Agda with Examples}
